Formalization of Forcing in Isabelle/ZF

### 2020-06-13 IJCAR2020/slides/forcing_IJCAR2020_slides.tex
minimal changes to the slides; draft mode for pictures

\begin{frame}
  \frametitle{Why Isabelle/ZF?}
  \begin{block}{Pros}
    \begin{itemize}
    \item Most advanced set theory formalized (around 2017).
    \item Closer to the mathematical practice of forcing.
    \item Structured proof language Isar
      \citep{DBLP:conf/tphol/Wenzel99}.
    \item Weak metatheory.
    \end{itemize}
  \end{block}
  %
  \pause
  %
  \begin{block}{Cons}
    \begin{itemize}
    \item A fraction of automation of Isabelle (\texttt{sledgehammer},
      etc).
    \item Untyped, and too weak a metatheory.
    \end{itemize}
  \end{block}
\end{frame}

@@

\begin{frame}
  \frametitle{Paulson's \texttt{ZF-Constructible}}

\end{frame}

@@

\begin{frame}
  \frametitle{The ctm approach to forcing, II}
  Given $G\sbq \PP$, we can adjoin it to $M$ to form the
  \textbf{generic extension} $M[G]$.
  \pause
  Every $a\in M[G]$ is coded by
  some $\punto{a}\in M$ through the definable (in $M$) function $\val$:
  \[
  M[G]\defi \{ \val(G,\punto{a}) : \punto{a}\in M\}
  \]
  \pause
  \emph{Fundamentally}, \textbf{truth} in $M[G]$ is coded in $M$ by the
  function $\forceisa$.
  \pause
  \begin{theorem}[Cohen, 1963]
    \begin{enumerate}
    \item $\bigl(\forall G\visible<5->{\alert{\ni p}}.\quad M[G], [\val(G,\punto{a})]
      \models \phi\bigr)
      \iff
      M,
      [\only<-4>{\1}\alert{\only<5->{p\hspace{0.2ex}}},\PP,\preceq,\punto{a}]\models
      \forceisa(\phi)$.
    \item $\forceisa$ is definable over $M$.\phantom{$\punto{a}).\models \phi\bigr)
      \iff M,$}\visible<6->{%
      $\overset{\searrow}{}$ \ 
      $\alert{p} \forces^M_{\PP,\preceq} \phi(\punto{a})$
      \  $\overset{\swarrow}{}$}
      \pause% 
      \pause% 
      \pause%       
    \item For $env\in\mathsf{list}(M)$,
      \[
      M[G], \mathrm{map}(\val(G),env) \models \phi)
      \iff
      \exists p\in G.\ M, [p,\preceq,\PP] \mathbin{{+}\!\!\!\!{+}} env \models
      \forceisa(\phi).
      \]
    \end{enumerate}
  \end{theorem} 
\end{frame}

\begin{frame}
  \frametitle{The ctm approach to forcing, III}
  By choosing the forcing notion $\lb \PP, {\preceq} ,\1\rb$
  appropriately one can tune the first order properties of  $M[G]$ (for
  any generic $G$).
  \pause

\end{frame}

@@

\begin{frame}
  \frametitle{What did we accomplish?}
  %
  \begin{shadowblock}{}
    \begin{enumerate}
    \item<+-> We adapted \texttt{ZF-Constructible} to obtain
      absoluteness results for nonempty transitive classes ($\leftarrow$
      \alert{in Isabelle2020}).
      %
      \pause

      Around 40 absoluteness/closure lemmas now hold using weaker
      hypotheses on the class $M$ (most of them, just that $M$ is
      transitive and nonempty). 
      %
      \pause
    \item<+-> We formalized the formula transformer $\forceisa$ and
      hence the forcing relation $\forces$, and proved the Fundamental
      Theorems.
    \item<+-> We showed that generic extensions of ctms of $\ZF$ are also
      ctms of $\ZF$ (respectively, adding $\AC$).
    \item<+-> We provided the forcing notion that adds a Cohen real,
      therefore proving the existence of a nontrivial extension.
    \end{enumerate}
  \end{shadowblock}
\end{frame}

\begin{frame}
  \frametitle{Some details}
  \begin{enumerate}
  \item<+-> Modifications to \texttt{ZF-Constructible}.

  \item<+-> We formalized the formula transformer $\forceisa$ and
    hence the forcing relation $\forces$, and proved the Fundamental
    Theorems.
  \item<+-> We showed that generic extensions of ctms of $\ZF$ are also
    ctms of $\ZF$ (respectively, adding $\AC$).
  \item<+-> We provided the forcing notion that adds a Cohen real,
    therefore proving the existence of a nontrivial extension.
  \end{enumerate}
\end{frame}

%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
\subsection{A map}

\begin{frame}
  \begin{center}
    \only<01>{\includegraphics[scale=0.23]{kunen_powerset_01.png} \\ \includegraphics[scale=0.26]{isa_converted_01.png}}%
    \only<02>{\includegraphics[scale=0.23]{kunen_powerset_02.png} \\ \includegraphics[scale=0.26]{isa_converted_02.png}}%
    \only<03>{\includegraphics[scale=0.23]{kunen_powerset_03.png} \\ \includegraphics[scale=0.26]{isa_converted_03.png}}%
    \only<04>{\includegraphics[scale=0.23]{kunen_powerset_04.png} \\ \includegraphics[scale=0.26]{isa_converted_04.png}}%
    \only<05>{\includegraphics[scale=0.23]{kunen_powerset_05.png} \\ \includegraphics[scale=0.26]{isa_converted_05.png}}%
    \only<06>{\includegraphics[scale=0.23]{kunen_powerset_06.png} \\ \includegraphics[scale=0.26]{isa_converted_06.png}}%
    \only<07>{\includegraphics[scale=0.23]{kunen_powerset_07.png} \\ \includegraphics[scale=0.26]{isa_converted_07.png}}%
    \only<08>{\includegraphics[scale=0.23]{kunen_powerset_08.png} \\ \includegraphics[scale=0.26]{isa_converted_08.png}}%
    \only<09>{\includegraphics[scale=0.23]{kunen_powerset_09.png} \\ \includegraphics[scale=0.26]{isa_converted_09.png}}%
    \only<10>{\includegraphics[scale=0.23]{kunen_powerset_10.png} \\ \includegraphics[scale=0.26]{isa_converted_10.png}}%
    \only<11>{\includegraphics[scale=0.23]{kunen_powerset_11.png} \\ \includegraphics[scale=0.26]{isa_converted_11.png}}%
    \only<12>{\includegraphics[scale=0.23]{kunen_powerset_12.png} \\ \includegraphics[scale=0.26]{isa_converted_12.png}}%
    \only<13>{\includegraphics[scale=0.23]{kunen_powerset_13.png} \\ \includegraphics[scale=0.26]{isa_converted_13.png}}%
    \only<14>{\includegraphics[scale=0.23]{kunen_powerset_14.png} \\ \includegraphics[scale=0.26]{isa_converted_14.png}}%
    \only<15>{\includegraphics[scale=0.23]{kunen_powerset_15.png} \\ \includegraphics[scale=0.26]{isa_converted_15.png}}%
    \only<16>{\includegraphics[scale=0.23]{kunen_powerset_16.png} \\ \includegraphics[scale=0.26]{isa_converted_16.png}}%
    \only<17>{\includegraphics[scale=0.23]{kunen_powerset_17.png} \\ \includegraphics[scale=0.26]{isa_converted_17.png}}%
    \only<18>{\includegraphics[scale=0.23]{kunen_powerset_18.png} \\ \includegraphics[scale=0.26]{isa_converted_18.png}}%
    \only<19>{\includegraphics[scale=0.23]{kunen_powerset_19.png} \\ \includegraphics[scale=0.26]{isa_converted_19.png}}%
    \only<20>{\includegraphics[scale=0.23]{kunen_powerset_20.png} \\ \includegraphics[scale=0.26]{isa_converted_20.png}}%
  \end{center}
\end{frame}

@@

%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
\subsection{}

\begin{frame}
  \frametitle{}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{}
%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
\subsection{}

\begin{frame}
  \frametitle{}
\end{frame}

@@

\begin{frame}
  \frametitle{Looking forward}
  \begin{block}{Formalizing math}
    \begin{itemize}
    \item<+-> Cofinality, K\H{o}nig's Theorem, Shanin's $\Delta$-system Lemma.
    \item<+-> Forcing notion for adding $\kappa$ Cohen reals.
    \item<+-> Theorems on preservation of cardinals.
    \end{itemize}
  \end{block}
  %
  \begin{block}{Technical aids}<+->
    \begin{itemize}
    \item<.-> Automatic relativization and proof of absoluteness of concepts.
    \item<+-> ``Relative functions'' (e.g., $\P^M$, $\card{\cdot}^M$, $\cf^M$).
    \item<+-> 
    \end{itemize}
  \end{block}
\end{frame}

>>>>

\begin{frame}
  \frametitle{Why Isabelle/ZF?}
  \begin{block}{Pros}
    \begin{itemize}
    \item Most advanced set theory formalized (around 2017).
      % \item Closer to the mathematical practice of forcing.
    \item Structured proof language Isar
      \citep{DBLP:conf/tphol/Wenzel99}.
    \item Weak metatheory.
    \end{itemize}
  \end{block}
  %
  \pause
  %
  \begin{block}{Cons}
    \begin{itemize}
    \item A fraction of automation of Isabelle (\texttt{sledgehammer},
      etc).
    \item Untyped, and too weak a metatheory.
    \end{itemize}
  \end{block}
\end{frame}

@@

\begin{frame}
  \frametitle{Paulson's \texttt{ZF-Constructible}}
  Let $C :: {}$ \isatt{i} $\Rightarrow$ \isatt{o} be a predicate on sets.
  \pause
  \begin{block}{Relativization of $\sbq$}
    $x \sbq^C y \equiv \forall z.\ C(z) \longrightarrow (z\in x
    \longrightarrow z\in y)$.
  \end{block}
  \pause
  \begin{block}{$\sbq$ is absolute}
    $\forall x\ y.\ C(x) \land C(y) \longrightarrow (x \sbq^C y
    \longleftrightarrow x\sbq y)$.
  \end{block}
  \pause 
  \texttt{ZF-Constructible} develops relativized version of the axioms
  and proves that many concepts are absolute for \textbf{transitive
    classes} $C$
  \[
  x\in y \land C(y)  \longrightarrow C(x).
  \]
  satisfying additional axioms.
\end{frame}

@@

\begin{frame}
  \frametitle{The ctm approach to forcing, II}
  Given $G\sbq \PP$, we can adjoin it to $M$ to form the
  \textbf{generic extension} $M[G]$.
  \pause
  Every $a\in M[G]$ is coded by
  some $\punto{a}\in M$ through the definable (in $M$) function $\val$:
  \[
  M[G]\defi \{ \val(G,\punto{a}) : \punto{a}\in M\}
  \]
  \emph{Fundamentally}, \textbf{truth} in $M[G]$ is coded in $M$ by the
  function $\forceisa$.
  \pause
  \begin{theorem}[\citep{Cohen-CH-PNAS}]
    \begin{enumerate}
    \item $\bigl(\forall G\visible<4->{\alert{\ni p}}.\quad M[G], [\val(G,\punto{a})]
      \models \phi\bigr)
      \iff
      M,
      [\only<-3>{\1}\alert{\only<4->{p\hspace{0.2ex}}},\PP,\preceq,\punto{a}]\models
      \forceisa(\phi)$.
    \item $\forceisa$ is definable over $M$.\phantom{$\punto{a}).\models \phi\bigr)
      \iff M,$}\visible<5->{%
      $\overset{\searrow}{}$ \ 
      $\alert{p} \forces^M_{\PP,\preceq} \phi(\punto{a})$
      \  $\overset{\swarrow}{}$}
      \pause% 
      \pause% 
      \pause%       
    \item For $env\in\mathsf{list}(M)$,
      \[
      M[G], \mathrm{map}(\val(G),env) \models \phi
      \iff
      \exists p\in G.\ M, [p,\preceq,\PP] \mathbin{{+}\!\!\!\!{+}} env \models
      \forceisa(\phi).
      \]
    \end{enumerate}
  \end{theorem} 
\end{frame}

\begin{frame}
  \frametitle{The ctm approach to forcing, III}
  By choosing the forcing notion $\lb \PP, {\preceq} ,\1\rb$
  appropriately one can tune the first order properties of  $M[G]$ (for
  any generic $G$).
  \pause
  %
  \begin{theorem}[\citep{Cohen-CH-PNAS}]
    If $\PP$ is the set of finite partial binary functions with domain
    included in $\aleph_2$, $M[G]$ satisfies the negation of the
    \textbf{Continuum Hypothesis}:
    \[
    M[G] \models 2^{\aleph_0} >\aleph_1
    \]
  \end{theorem}
\end{frame}

@@

\begin{frame}
  \frametitle{What did we accomplish?}
  %
  \begin{shadowblock}{}
    \begin{enumerate}
    \item<1-> We adapted \texttt{ZF-Constructible} to obtain
      absoluteness results for nonempty transitive classes ($\leftarrow$
      \alert{in Isabelle2020}).

      \visible<2->{
        Around 40 absoluteness/closure lemmas now hold using weaker
        hypotheses on the class $C$ (most of them, just that $C$ is
        transitive and nonempty).
      }
    \item<3-> We formalized the formula transformer $\forceisa$ and
      hence the forcing relation $\forces$, and proved the Fundamental
      Theorems.
    \item<4-> We showed that generic extensions of ctms of $\ZF$ are also
      ctms of $\ZF$ (respectively, adding $\AC$).
    \item<5-> We provided the forcing notion that adds a Cohen real,
      therefore proving the existence of a nontrivial extension.
    \end{enumerate}
  \end{shadowblock}
\end{frame}

\begin{frame}
  \frametitle{Some details}
  \begin{enumerate}\setcounter{enumi}{1}
  \item<+-> We formalized the formula transformer $\forceisa$ and
    hence the forcing relation $\forces$, and proved the Fundamental
    Theorems.
    \begin{itemize}
    \item<+-> 
    \item<+-> 
    \end{itemize}
  \item<+-> We showed that generic extensions of ctms of $\ZF$ are also
    ctms of $\ZF$ (respectively, adding $\AC$).
    \begin{itemize}
    \item<+-> We can see a bit of the code, compared to the actual math.
    \end{itemize}
  \end{enumerate}
\end{frame}

%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
\subsection{Math vs Code}

\begin{frame}
  \begin{center}
    \only<01>{\includegraphics[draft,scale=0.23]{kunen_powerset_01.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_01.png}}%
    \only<02>{\includegraphics[draft,scale=0.23]{kunen_powerset_02.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_02.png}}%
    \only<03>{\includegraphics[draft,scale=0.23]{kunen_powerset_03.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_03.png}}%
    \only<04>{\includegraphics[draft,scale=0.23]{kunen_powerset_04.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_04.png}}%
    \only<05>{\includegraphics[draft,scale=0.23]{kunen_powerset_05.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_05.png}}%
    \only<06>{\includegraphics[draft,scale=0.23]{kunen_powerset_06.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_06.png}}%
    \only<07>{\includegraphics[draft,scale=0.23]{kunen_powerset_07.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_07.png}}%
    \only<08>{\includegraphics[draft,scale=0.23]{kunen_powerset_08.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_08.png}}%
    \only<09>{\includegraphics[draft,scale=0.23]{kunen_powerset_09.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_09.png}}%
    \only<10>{\includegraphics[draft,scale=0.23]{kunen_powerset_10.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_10.png}}%
    \only<11>{\includegraphics[draft,scale=0.23]{kunen_powerset_11.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_11.png}}%
    \only<12>{\includegraphics[draft,scale=0.23]{kunen_powerset_12.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_12.png}}%
    \only<13>{\includegraphics[draft,scale=0.23]{kunen_powerset_13.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_13.png}}%
    \only<14>{\includegraphics[draft,scale=0.23]{kunen_powerset_14.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_14.png}}%
    \only<15>{\includegraphics[draft,scale=0.23]{kunen_powerset_15.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_15.png}}%
    \only<16>{\includegraphics[draft,scale=0.23]{kunen_powerset_16.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_16.png}}%
    \only<17>{\includegraphics[draft,scale=0.23]{kunen_powerset_17.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_17.png}}%
    \only<18>{\includegraphics[draft,scale=0.23]{kunen_powerset_18.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_18.png}}%
    \only<19>{\includegraphics[draft,scale=0.23]{kunen_powerset_19.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_19.png}}%
    \only<20>{\includegraphics[draft,scale=0.23]{kunen_powerset_20.png} \\ \includegraphics[draft,scale=0.26]{isa_converted_20.png}}%
  \end{center}
\end{frame}

@@

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% \section{}
%% %-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
%% \subsection{}
%% 
%% \begin{frame}
%%   \frametitle{}
%% \end{frame}

@@

\begin{frame}
  \frametitle{Looking forward}
  \begin{block}{Formalizing math}
    \begin{itemize}
    \item<+-> Cofinality, K\H{o}nig's Theorem, Shanin's $\Delta$-system Lemma.
    \item<+-> Forcing notion for adding $\kappa$ Cohen reals.
    \item<+-> Theorems on preservation of cardinals.
    \end{itemize}
  \end{block}
  %
  \begin{block}{Technical aids}<+->
    \begin{itemize}
    \item<.-> Automatic relativization and proof of absoluteness of concepts.
    \item<+-> ``Relative functions'' (e.g., $\P^M$, $\card{\cdot}^M$, $\cf^M$).
    \end{itemize}
  \end{block}
\end{frame}
